---
title: "Attribute Wise Learning for Scoring Outliers"
author: "Danny Morris"
date: "July 18, 2018"
output: 
    html_document:
        theme: lumen
        toc: true
        df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      cache = TRUE)
```

# Overview

Attirubte-wise Learning for SCoring Outliers (ALSO) is an outlier detection technique for standard multidimensional datasets. For each variable in the dataset, a predictive model is constructed in which the remaining features are predictors. When a feature is generally predictable with good accuracy, observations deviating significantly from predicted values may be outliers. Observations with significant deviations across many predictable features may be strong outliers.

This document will illustrate the use of the **ALSO package**. It was created for the sole purpose of making the ALSO technique easily available to R users. The package currently contains a single function, `ALSO()`.

```{r}
# devtools::install_github("dannymorris/ALSO")
library(ALSO)

library(dplyr)
```

# Data

```{r}
states <- state.x77 %>% as_tibble()

states
```
